A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-29 latex/main.tex

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed a more comprehensive mutational landscape
of SARS-CoV-2 by mapping deletions onto the ARG
using post-hoc parsimony.
We identified important deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur in major VOCs
(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
\cite{Akaishi2023,Li2023indels}).
Eight of these deletions occur in S1-NTD,
a known hotspot for recurrent deletions \cite{McCarthy2021}.

[This is a separate paramgraph discussing where the model has 
broken down a bit]
There are three contiguous deletions in NSP6
(4-base at 11288, 5-base at 11283, and 5-base at 11292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node and
the other above the BA.2 origin node in the ARG
(Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up
during deletion remapping because they had
overlapping four sites at 11288-11292
(which was mapped as the 4-base deletion event).

Furthermore, the placement of these [Which events? All of them? just 
the ones we didn't properly place?] deletion events in the ARG
reflects that of the VOC-associated deletion events
inferred in a previous study \cite{Akaishi2023}.

We found highly recurrent deletions using the ARG,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG
shown in Table~\ref{tab:rec_dels}.
About half of these deletions are one-base and
occur in the 5' or 3' UTR of the SARS-CoV-2 genome,
which are enriched for poor-quality sequencing data.
Genuine recurrent deletions are also captured in the ARG,
for example, two known recurrent deletions in S1-NTD
associated with antibody escape \cite{McCarthy2021}.
In-depth analysis of the deletions in the ARG
is beyond the scope of this study,
but we believe that the ARG will greatly
facilitate future evolutionary studies of
deletions in SARS-CoV-2.

@@

\subsubsection{XBH}
All the XBH samples descend from a well supported recombination node,
despite two reversions (T222C and T12789C, respectively),
which occur at excluded sites and
were remapped by parsimony during postprocessing.
If these two sites had been used during ARG inference,
the branch attached to BA.2.1 would have been attached to
the BA.2.5 shown on the top left instead,
leading to a more parsimonious local topology
with two mutations on the left branch.

\subsubsection{XBQ/XBK/XBK.1/CJ.1.3}
https://github.com/jeromekelleher/sc2ts-paper/issues/349
https://github.com/jeromekelleher/sc2ts-paper/issues/350

>>>>

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed a more comprehensive mutational landscape
of SARS-CoV-2 by mapping deletions onto the ARG
using post-hoc parsimony.
We identified major deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur in major VOCs
(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
\cite{Akaishi2023,Li2023indels}).
Eight of these deletions occur in S1-NTD,
a known hotspot for recurrent deletions \cite{McCarthy2021}.

The placement of the following major deletion events
in the ARG matches the phylogenetic placement of
VOC-associated deletion events
suggested in a previous study \cite{Akaishi2023}:
a 3-base deletion 21,991 above the Alpha origin node;
two 6-base deletions at 22,029 and 28,248 above the Delta origin node;
a 9-base deletion at 28,362 above Omicron (B.1.1.529) origin node;
a 3-base deletion at 6,513,
a 9-base deletion at 21,987, and
a 3-base deletion at 22,194 above the BA.1 origin node;
a 9-base deletion at 21,633 above the BA.2 origin node.

However, the placement of several major deletion events
in the ARG differs from the phylogenetic placement of
the corresponding deletion events suggested
in Akaishi and Fujiwara \cite{Akaishi2023}.
That study \cite{Akaishi2023} suggests that
a 6-base deletion at 21,765 occurred
before the divergence of Alpha and Omicron,
but in the ARG this deletion appears to be recurrent,
occurring above the origin nodes of Alpha, BA.1, and BA.4.
Similarly, that study \cite{Akaishi2023} suggests that
a 1-base deletion at 28,271 occurred
before the divergence of Alpha, Delta, and Omicron,
but it is mapped as a recurrent deletion in the ARG
(at the Alpha origin node and above node 1436808).
These differences suggest that
some key deletions carry important phylogenetic signals
ignored by the HMM used by sc2ts.

% This is a separate paragraph discussing where the model has 
% broken down a bit.
Furthermore, there are three contiguous deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node (starting at 11,283) and
the other above the BA.2 origin node (starting at 11,288)
in the ARG (Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up during deletion remapping
because they had four overlapping sites at 11,288-11,292
(which was mapped as the 4-base deletion event).
Furthermore, the analysis of
Akaishi and Fujiwara \cite{Akaishi2023} suggests that
a 9-base deletion at 11,288
occurred before the divergence of Alpha and Omicron.
In the ARG, however, this deletion is mapped
above the Alpha origin node.

By mapping deletion events on the ARG,
we also identified highly recurrent deletions,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG
shown in Table~\ref{tab:rec_dels}.
About half of these deletions are 1-base and
occur in the 5' or 3' UTR of the SARS-CoV-2 genome
(from the start to 250 and from 29,700 to the end),
which are enriched for poor-quality sequencing data.
Genuine recurrent deletions are captured in the ARG,
for example, two known recurrent deletions in S1-NTD
associated with antibody escape \cite{McCarthy2021}.
There are also 1-base recurrent deletions
at 29,541, 29,543, and 29,555
in a non-coding region (between N and ORF10)
which are plausibly homoplasic as they might arise due to
polymerase skipping during genome replication and
this regions is likely free from strong selective pressure.
An in-depth analysis of all the recurrent deletions in the ARG
is beyond the scope of this study,
but we believe that the ARG will serve as a rich resource
to facilitate future evolutionary studies of
deletions in SARS-CoV-2.

@@

\subsubsection{XBH}
All the XBH samples descend from a well supported recombination node,
despite two reversions (T222C and T12789C, respectively),
which occur at excluded sites and
were remapped by parsimony during postprocessing.
If these two sites had been used during ARG inference,
the branch attached to BA.2.1 would have been attached to
the BA.2.5 shown on the top left instead,
leading to a more parsimonious local topology
with two mutations on the left branch. This would also remove the two
blue lines on the left and side of the copying pattern.

\subsubsection{XBQ/XBK/XBK.1}
https://github.com/jeromekelleher/sc2ts-paper/issues/349
https://github.com/jeromekelleher/sc2ts-paper/issues/350
